---
title: 'compgen2021: Week 1 exercises'
author: "O. Daniel López Olmos"
output:
  html_document:
    df_print: paged
  pdf: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercises for Week1

## Statistics for genomics 

### How to summarize collection of data points: The idea behind statistical distributions

1. Calculate the means and variances 
of the rows of the following simulated data set, and plot the distributions
of means and variances using `hist()` and `boxplot()` functions. [Difficulty: **Beginner/Intermediate**]  
```{r getDataChp3Ex}
#import libraries
library(mosaic, warn.conflicts = F)

set.seed(100)

#sample data matrix from normal distribution
gset=rnorm(600,mean=200,sd=70)
data=matrix(gset,ncol=6)

#calculate means and variances 
rowMeans <- apply(data, 1, mean)
rowVars <- apply(data, 1, var)

#plot the distributions
hist(rowMeans, col = "lightblue", border = "white",
     main = "Distribution of the means calculated row-wise",
     xlab = "Row means")

hist(rowVars, col = "lightblue", border = "white",
     main = "Distribution of the variances calculated row-wise",
     xlab = "Row variances")

boxplot(rowMeans, col = "lightblue", border = "white",
     main = "Distribution of the means")

boxplot(rowVars, col = "lightblue", border = "white",
     main = "Distribution of the variances")
```



2. Using the data generated above, calculate the standard deviation of the
distribution of the means using the `sd()` function. Compare that to the expected
standard error obtained from the central limit theorem keeping in mind the
population parameters were  $\sigma=70$ and $n=6$. How does the estimate from the random samples change if we simulate more data with
`data=matrix(rnorm(6000,mean=200,sd=70),ncol=6)`? [Difficulty: **Beginner/Intermediate**] 
```{r}


```


3. Simulate 30 random variables using the `rpois()` function. Do this 1000 times and calculate the mean of each sample. Plot the sampling distributions of the means
using a histogram. Get the 2.5th and 97.5th percentiles of the
distribution. [Difficulty: **Beginner/Intermediate**] 
```{r exRpoisChp3}
require(mosaic, quietly = T)
#HINT
set.seed(100)

#sample 30 values from poisson dist with lambda paramater = 30
pois1 <- rpois(30, lambda=5)
#bootstrap resampling
bootMeans = mosaic::do(1000) * mean(resample(pois1))
#get percentiles from the bootstrap means
q = quantile(bootMeans[, 1], p = c(0.025, 0.975))

#plot sampling distributions
hist(bootMeans[,1], col = "lightblue", border = "white",
                    xlab = "Sample means")
abline(v = c(q[1], q[2] ), col = "black")
text(x = q[1], y = 200, round(q[1], 3), adj = c(1, 0))
text(x = q[2], y = 200, round(q[2], 3), adj = c(0, 0))
```
4. Use the `t.test()` function to calculate confidence intervals
of the mean on the first random sample `pois1` simulated from the `rpois()` function below. [Difficulty: **Intermediate**] 
```{r}
#calculate confidence intervals with 't.test()'
t.test(pois1)
```

Confidence intervals: [4.362104, 5.504563]

5. Use the bootstrap confidence interval for the mean on `pois1`. [Difficulty: **Intermediate/Advanced**] 
```{r}

```



### How to test for differences in samples

1. Test the difference of means of the following simulated genes
using the randomization, `t-test()`, and `wilcox.test()` functions.
Plot the distributions using histograms and boxplots. [Difficulty: **Intermediate/Advanced**] 
```{r exRnorm1chp3}

set.seed(101)
gene1 <- rnorm(30, mean = 4, sd = 3)
gene2 <- rnorm(30, mean = 3, sd = 3)
#calculate original difference
orig_diff <- mean(gene1) - mean(gene2)


#create a df for our data
genesDF <- data.frame(exp = c(gene1, gene2), group = c(rep("group1", 30), rep("group2", 30)))
#null distribution-no difference in means 
exp_null <- do(1000) * diff(mosaic::mean(exp ~ shuffle(group), data = genesDF))

#test for differences using randomization
hist(exp_null[, 1], xlab = "Null distribution | no difference in samples",
     main = expression(paste(H[0]," :no difference in means") ),
     xlim = c(-2,2), col = "lightblue", border = "white")
abline(v = quantile(exp_null[, 1], 0.95), col = "black")
abline(v = orig_diff, col = "red" )
text(x = quantile(exp_null[, 1], 0.95), y = 200, "0.05", adj = c(1, 0), col = "black")
text(x = orig_diff, y = 250, "orig_diff.", adj = c(1, 0), col = "red")

```

Test the difference of means of the simulated genes using `t-test()`.
```{r}
#test differences of means using 't.test()'
stats::t.test(gene1, gene2)
```

Test the difference of means of the simulated genes using `wilcox.test()`.
```{r}
wilcox.test(gene1, gene2, conf.int = T)
```

Plot the distributions using histograms and boxplots.
```{r}
#plot the distributions of gene expressions
hist(genesDF[, 1], #histogram
     main = "Distribution of gene expressions",
     xlab = "Gene expression",
     col = "lightblue", border = "white")
#boxplot
boxplot(genesDF[,1],
     main = "Boxplot of gene expressions",
     col = "lightblue", border = "black")
```


2. Test the difference of the means of the following simulated genes
using the randomization, `t-test()` and `wilcox.test()` functions.
Plot the distributions using histograms and boxplots. [Difficulty: **Intermediate/Advanced**] 
```{r exRnorm2chp3}

set.seed(100)
gene1 = rnorm(30, mean = 4, sd = 2)
gene2 = rnorm(30, mean = 2, sd = 2)
#calculate original difference
orig_diff = mean(gene1) - mean(gene2)


#create a df for our data
genesDF = data.frame(exp = c(gene1, gene2), group = c(rep("group1", 30), rep("group2", 30)))
#null distribution-no difference in means 
exp_null <- do(1000) * diff(mosaic::mean(exp ~ shuffle(group), data = genesDF))

#test for differences using randomization
hist(exp_null[, 1], xlab = "Null distribution | no difference in samples",
     main = expression(paste(H[0]," :no difference in means") ),
     xlim = c(-2,2), col = "lightblue", border = "white")
abline(v = quantile(exp_null[, 1], 0.95), col = "black")
abline(v = orig_diff, col = "red" )
text(x = quantile(exp_null[, 1], 0.95), y = 200, "0.05", adj = c(1, 0), col = "black")
text(x = orig_diff, y = 250, "orig_diff.", adj = c(1, 0), col = "red")
```

Test the difference of means of the simulated genes using `t-test()`.
```{r}
#test differences of means using 't.test()'
stats::t.test(gene1, gene2)
```

Test the difference of means of the simulated genes using `wilcox.test()`.
```{r}
wilcox.test(gene1, gene2, conf.int = T)
```

Plot the distributions using histograms and boxplots.
```{r}
#plot the distributions of gene expressions
hist(genesDF[, 1], #histogram
     main = "Distribution of gene expressions",
     xlab = "Gene expression",
     col = "lightblue", border = "white")
#boxplot
boxplot(genesDF[,1],
     main = "Boxplot of gene expressions",
     col = "lightblue", border = "black")
```


3. We need an extra data set for this exercise. Read the gene expression data set as follows:
`gexpFile=system.file("extdata","geneExpMat.rds",package="compGenomRData") data=readRDS(gexpFile)`. The data has 100 differentially expressed genes. The first 3 columns are the test samples, and the last 3 are the control samples. Do 
a t-test for each gene (each row is a gene), and record the p-values.
Then, do a moderated t-test, as shown in section "Moderated t-tests" in this chapter, and record 
the p-values. Make a p-value histogram and compare two approaches in terms of the number of significant tests with the $0.05$ threshold.
On the p-values use FDR (BH), Bonferroni and q-value adjustment methods.
Calculate how many adjusted p-values are below 0.05 for each approach.
[Difficulty: **Intermediate/Advanced**] 
```{r}
require(matrixStats, quietly = T)
#import data
gexpFile = system.file("extdata", "geneExpMat.rds", package = "compGenomRData") 
data = readRDS(gexpFile)

#true difference in means
dx = rowMeans(data[, 1:3]) - rowMeans(data[, 4:6])

n1 = 3
n2 = 3
#get the esimate of pooled variance 
stderr = sqrt((rowVars(data[, 1:3])*(n1-1) + rowVars(data[, 4:6])*(n2-1)) / (n1+n2-2) * ( 1/n1 + 1/n2 ))
#do the shrinking towards median
mod.stderr = (stderr + median(stderr)) / 2 # moderation in variation

#do a t-test for each gene and record p-values
t = dx / stderr
p = 2*pt( -abs(t), n1+n2-2 )

#do a moderated t-test and record p-values
t.mod <- dx / mod.stderr
p.mod = 2*pt( -abs(t.mod), n1+n2-2 )

#make a p-value histogram and compare both approaches
par(mfrow = c(1, 2))

hist(p, col = "lightblue", border = "white", main = "",
     xlab = "P-values t-test")
mtext(paste("Signifcant tests:", sum(p < 0.05))  )

hist(p.mod, col = "lightblue", border="white", main="",
     xlab="P-values mod. t-test")
mtext(paste("Signifcant tests:", sum(p.mod <0.05)))
```

On the p-values use FDR (BH), Bonferroni and q-value adjustment methods.
Calculate how many adjusted p-values are below 0.05 for each approach.
```{r}
require(qvalue, quietly = T)

#FDR adjusted value
fdr.adj.pval = p.adjust(p, method = "fdr")
#Bonferroni correction
bonf.pval=p.adjust(p, method = "bonferroni")
#q-value
qvalues <- qvalue(p)$q

n_pvalues <- data.frame(fdr_adj_pval = c(sum(fdr.adj.pval < 0.05)),
                        bonf_pval = sum(bonf.pval < 0.05),
                        q_values = sum(qvalues < 0.05))
rownames(n_pvalues) <- "# of p-values below 0.05"
n_pvalues
```


### Relationship between variables: Linear models and correlation

Below we are going to simulate X and Y values that are needed for the 
rest of the exercise.
```{r exLM1chp3}
# set random number seed, so that the random numbers from the text
# is the same when you run the code.
set.seed(32)

# get 50 X values between 1 and 100
x = runif(50, 1, 100)

# set b0,b1 and variance (sigma)
b0 = 10
b1 = 2
sigma = 20
# simulate error terms from normal distribution
eps = rnorm(50, 0, sigma)
# get y values from the linear equation and addition of error terms
y = b0 + b1*x+ eps
```


1. Run the code then fit a line to predict Y based on X. [Difficulty:**Intermediate**] 
```{r}
#instance the model
model <- lm(y ~ x)
```


2. Plot the scatter plot and the fitted line. [Difficulty:**Intermediate**] 
```{r}
#scatter plot
plot(x, y, pch=20)
#plot the fitted line
abline(model, col = "red")
```


3. Calculate correlation and R^2. [Difficulty:**Intermediate**] 
```{r}
cat("Correlation of model: ", correlation <- cor(x, y), "\n")
R2 <-summary.lm(model)$r.squared
cat("R squared of model: ", R2)
```

Correlation shows a high value of correlation between both variables.
The R^2 is sufficiently good to assume that the 'x' variable has the power to predict 'y' values.

4. Run the `summary()` function and 
try to extract P-values for the model from the object
returned by `summary`. See `?summary.lm`. [Difficulty:**Intermediate/Advanced**] 
```{r}
#p-value: < 2.2e-16
sum <- summary.lm(model)
sum
```

5. Plot the residuals vs. the fitted values plot, by calling the `plot()` 
function with `which=1` as the second argument. First argument
is the model returned by `lm()`. [Difficulty:**Advanced**] 
```{r}
plot(model, which = 1)
```

The plot above shows no obvious pattern between our residuals and fitted values, showing no tendecy in our model.

6. For the next exercises, read the data set histone modification data set. Use the following to get the path to the file:
```{r}
#import data
hmodFile=system.file("extdata",
                    "HistoneModeVSgeneExp.rds",
                     package="compGenomRData")
data2 <- readRDS(hmodFile)
```


There are 3 columns in the dataset. These are measured levels of H3K4me3,
H3K27me3 and gene expression per gene. Once you read in the data, plot the scatter plot for H3K4me3 vs. expression. [Difficulty:**Beginner**] 
```{r}
library(ggplot2)

#plot H3K4me3 vs. expression.
ggplot(data2, aes(x = H3k4me3, y = measured_log2)) + 
  geom_point()
```


7. Plot the scatter plot for H3K27me3 vs. expression. [Difficulty:**Beginner**] 
```{r}
#plot H3K27me3 vs. expression
ggplot(data2, aes(x = H3k27me3, y = measured_log2)) + 
  geom_point()
```

8. Fit the model for prediction of expression data using: 1) Only H3K4me3 as explanatory variable, 2) Only H3K27me3 as explanatory variable, and 3) Using both H3K4me3 and H3K27me3 as explanatory variables. Inspect the `summary()` function output in each case, which terms are significant. [Difficulty:**Beginner/Intermediate**] 
1)
```{r}
#fitting a regression model using H3K4me3 as explanatory variable
model2 <- lm(measured_log2 ~ H3k4me3, data = data2)
summary(model2)
```

From the above output, the first thing to notice is the p-value of the model, which is close to 0, meaning that we can trust the model.
As so, p-values for the explanatory variable 'H3K4me3' and for the intercept are really trustworthy.

2)
```{r}
#fitting a regression model using H3K27me3 as explanatory variable
model3 <- lm(measured_log2 ~ H3k27me3, data = data2)
summary(model3)
```

Our second model, has the same p-values for the explanatory variable 'H3K27me3', intercept and overall model. Nevertheless, it's R-squared score is less significant that the score of our previous model (0.04449 < 0.6511), which indicates that the correlation between 'H3K27me3' and gene expression 'measured_log2' is stronger than 'H3K27me3' and gene expression.
3)
```{r}
#fitting a regression model using both epigenetic marks as explanatory variables
model4 <- lm(measured_log2 ~ H3k4me3 + H3k27me3, data = data2)
summary(model4)
```

Our multivariate model shows to be a better model that the models using a single explanatory variable, with a R-squared score =  0.6723. Which suggests that predicting gene expression is better when both epigenetic marks are used in modeling.

10. Is using H3K4me3 and H3K27me3 better than the model with only H3K4me3? [Difficulty:**Intermediate**] 

Both summary statistics shows that both models are competent in predicting gene expression. For a better answer, let's compare the correlation coefficient.
```{r}
singlemodelCorr <- summary.lm(model2)$r.squared
multimodelCorr <- summary.lm(model4)$r.squared

cat("Correlation of model with only H3K4me3 as explanatory variable: ", singlemodelCorr, "\n")

cat("Correlation of model with H3K4me3 and H3K27me3 as explanatory variables: ", multimodelCorr)
```

We can appreciate a slight increase in the correlation coefficient of the model that uses both epigenetic marks, so we can conclude that this model is better than the model using just H3K4me3.

