---
title: 'compgen2021: Week 3 exercises'
author: 'O. Daniel López Olmos'
output:
  pdf_document: default
  pdf: default
---

# Exercises for Week 3


### Classification 
For this set of exercises we will be using the gene expression and patient annotation data from the glioblastoma patient. You can read the data as shown below:
```{r,readMLdataEx}
library(compGenomRData)
# get file paths
fileLGGexp = system.file("extdata",
                      "LGGrnaseq.rds",
                      package="compGenomRData")
fileLGGann = system.file("extdata",
                      "patient2LGGsubtypes.rds",
                      package="compGenomRData")

# gene expression values
geneExp <- readRDS(fileLGGexp)
# transpose data in order to have features on the columns
geneExp <- t(geneExp)

# patient annotation
patient <- readRDS(fileLGGann)

# merge the data
geneExp <- merge(patient, geneExp, by = "row.names" )
# set sample ids
rownames(geneExp) = geneExp[, 1]
geneExp <- geneExp[, -1]
```

1. Our first task is to not use any data transformation and do classification. Run the k-NN classifier on the data without any transformation or scaling. What is the effect on classification accuracy for k-NN predicting the CIMP and noCIMP status of the patient? [Difficulty: **Beginner**]

**solution:**

Giving the example in the supervised learning chapter, here for both the training and testing predictions, the accuracy scores had a decrease(as expected due to no data preprocessing). For the training data, accuracy score dropped down ~10% and for the testing data ~13%.
Note: here we dividied tha data
```{r} 
require(caret, quietly = T )
set.seed(1212)

data_idx <- createDataPartition(y = geneExp[,1], p = 0.7, )[[1]]
trainingData <- geneExp[data_idx,]
x_train <- trainingData[, -1]
y_train <- trainingData[, 1]

testData <- geneExp[-data_idx,]
x_test <- testData[, -1]
y_test <- testData[, 1]

#instance and fit the model
KNNfit <- knn3(x = x_train, y = y_train, 
                 k = 5)

# PREDICT DE CARET O DE STATS
#predict for training data
trainPreds <- predict(KNNfit, x_train, type = "class")

#get confusion matrix
confusionMatrix(data = y_train, reference = trainPreds)

```


```{r}
#predict for testing data
testPreds <- predict(KNNfit, x_test, type = "class")

#confusion matrix
confusionMatrix(data = y_test, reference = testPreds)
```


2. Bootstrap resampling can be used to measure the variability of the prediction error. Use bootstrap resampling with k-NN for the prediction accuracy. How different is it from cross-validation for different $k$s? [Difficulty: **Intermediate**]
**solution:**
We will use the caret function 'trainControl()' to set a resampling method and test boostrap and cross-validation with k=3,5,7 which are the most common k values.
For this excercise, cross-validation strategy gives slightly better accuracy score, improving from ~75% with bootstrap to ~77% CV.
note: Here we have only used the half of the total variables in the data, since R has some problems when making opperations with a large number of variables(this can be fixed changing the ppsize option in our .Rprofile file) so, the actual prediction accuracy could sustantally be improved whe using all variables. 
```{r}
set.seed(400)

#split data variables in half
geneExp <- geneExp[,1:10251]
data_idx <- createDataPartition(y = geneExp[,1], p = 0.7, )[[1]]
trainingData <- geneExp[data_idx,]
x_train <- trainingData[, -1]
y_train <- trainingData[, 1]

testData <- geneExp[-data_idx,]
x_test <- testData[, -1]
y_test <- testData[, 1]
#bootstrap resampling
#define training control
train_control <- trainControl(method = "boot", number = 10)
#train the model
bootModel <- train(subtype ~., data = trainingData, method = "knn",
               trControl = train_control, 
               tuneGrid = data.frame(k = c(3,5,7)))

#plot prediction accuracy for each k
plot(x = c(3,5,7), bootModel$results[, 2] * 100,
     main = "Prediction accuracy using Bootstrap for k=3,5,7",
     pch = 19,
     ylab = "Prediction acuracy", xlab = "k")


```

```{r}
#cross validation with k=3,5,7
set.seed(406)

#define training control
crossValcontrol <- trainControl(method = "cv", number = 10)
#train the model
crossValmodel <- train(subtype ~., data = trainingData, method = "knn",
               trControl = crossValcontrol, 
               tuneGrid = data.frame(k = c(3,5,7)))

#plot prediction error variability for each k
plot(x = c(3,5,7), crossValmodel$results[, 2] * 100,
     main = "Prediction accuracy using CV with k=3,5,7",pch = 19,
     ylab = "Prediction accuracy", xlab = "k")

```


3. There are a number of ways to get variable importance for a classification problem. Run random forests on the classification problem above. Compare the variable importance metrics from random forest and the one obtained from DALEX. How many variables are the same in the top 10? [Difficulty: **Advanced**]
**solution:**
```{r}
library(ranger)
set.seed(735)

ctlForest <- trainControl(method = "none", 
                          classProbs = T,
                          savePredictions = T)
#instance model
rforestModel <- train(subtype ~.,
                      method = "ranger",
                      data = trainingData,
                      trControl = ctlForest,
                      importance = "permutation", #permutation drop-out to compare with DALEX
                      tuneGrid = data.frame(mtry = 10, min.node.size = 1, 
                                            splitrule = "gini"))
#plot the 10 most important variables
plot(varImp(rforestModel), top = 10)
```


```{r}
require(DALEX, quietly = T)
#variable importance with DALEX


set.seed(427)
#permutation drop-out
featImportance <- DALEX::explain(rforestModel, 
                               label = "random forest", 
                               data = trainingData[,-1], 
                               #transform y to numerical with 1=CIMP and 2=noCIMP
                               y = as.numeric(trainingData[,1]))

visFeatImp <- feature_importance(featImportance, n_sample = 10,
                                 #get a single-permutation-based measure
                                 B = 1,
                                 type = "difference")
plot(visFeatImp)

```


4. Come up with a unified importance score by normalizing importance scores from random forests and DALEX, followed by taking the average of those scores. [Difficulty: **Advanced**]
**solution:**
```{r}
require(BBmisc, quietly = T)

#normalize importance scores from random forests
impForest <- varImp(rforestModel)
impForest <- BBmisc::normalize(impForest$importance)
#normalize importance scores from DALEX
impDalex <- explain(rforestModel, 
                    label = "random forest", 
                    data = x_train, 
                    y = as.numeric(trainingData[,1]))

impDalex <- normalize(impDalex)
#taking the average of the scores
#COMPARAR/CAMBIAR CON UNA FUNCION APPLY
avgForest <- mean(impForest)
avgDalex <- mean(impDalex)
#create a df showing both scores
impScore <- data.frame(random.forest.score = avgForest,
                       DALEX.score = avgDalex)  

head(impScore)
```



### Regression
For this set of problems we will use the regression data set where we tried to predict the age of the sample from the methylation values. The data can be loaded as shown below: 
```{r, readMethAgeex}

# file path for CpG methylation and age
fileMethAge <- system.file("extdata",
                      "CpGmeth2Age.rds",
                      package = "compGenomRData")

# read methylation-age table
ageMeth <- readRDS(fileMethAge)
```

1. Run random forest regression and plot the importance metrics. [Difficulty: **Beginner**]

**solution:**
```{r}
set.seed(18)
par(mfrow=c(1,2))

#use only half of the variables
ageMeth <- ageMeth[,1:13789]
#remove variables with low variation
ageMeth <- ageMeth[, c(TRUE, matrixStats::colSds(as.matrix(ageMeth[,-1]))>0.1)]
data_idx2 <- createDataPartition(y = ageMeth[,1], p = 0.7, )[[1]]
trainingData2 <- ageMeth[data_idx2,]

testData2 <- ageMeth[-data_idx2,]

trctrl <- trainControl(method = "none")

#train the model
rfRegModel <- train(Age~., 
               data = trainingData2, 
               method = "ranger",
               trControl =trctrl,
               # calculate importance
               importance = "permutation", 
               tuneGrid = data.frame(mtry = 50,
                                     min.node.size = 5,
                                     splitrule = "variance")
               )


plot(trainingData2$Age, rfRegModel$finalModel$predictions,
     pch = 19, xlab = "Observed Age",
     ylab = "OOB predicted Age")
mtext(paste("R-squared",
            format(rfRegModel$finalModel$r.squared, digits = 2)))

# plot residuals
plot(trainingData2$Age, (rfRegModel$finalModel$predictions -                    trainingData2$Age),
     pch = 18, ylab = "Residuals (predicted-observed)",
     xlab = "Observed Age", col = "blue3")
abline(h = 0, col = "red4", lty = 2)
```


2. Split 20% of the methylation-age data as test data and run elastic net regression on the training portion to tune parameters and test it on the test portion. [Difficulty: **Intermediate**] 

**solution:**
```{r}
library(glmnet)
set.seed(848)

#split the data
dataIdx <- createDataPartition(y = ageMeth[,1], p = 0.8, )[[1]] #index for 80% of the data
trainData <- ageMeth[dataIdx,]
X_train <- trainData[, -1]
Y_train <- trainData[, 1]

testData <- ageMeth[-dataIdx,]
X_test <- testData[, -1]
Y_test <- testData[, 1]

#do cross validation so we can get the best parameters
ctrlNet <- trainControl(method = "cv", number = 10)

#train the model
netModel <- train(Age ~ ., data = trainData, 
                 method = "glmnet",
                 trControl = ctrlNet,
                 # alpha and lambda parameters to try
                 tuneGrid = data.frame(alpha = seq(0.1, 0.3, 0.5),
                                       lambda = seq(0.1,0.7,0.05)))

#best parameter values by cross-validation accuracy
netModel$bestTune
```

```{r}
#CHECAR SI EL PARAMETRO 'TRCONTROL' SE PUEDE USAR EL MODELO ANTERIOR
#update the model with best parameters suggested
UpdatedNetModel <- train(Age ~ ., data = trainData, 
                 method = "glmnet",
                 
                 trControl = trainControl("none"),
                 #set alpha and lambda as suggested
                 tuneGrid = data.frame(alpha = 0.1,
                                       lambda = 0.7))
#predict using test data
netPredictions <-predict.glmnet(UpdatedNetModel, X_test)
#show accuracy
#confusion.glmnet(netPredictions, newy = Y_test)
class.res=predict(enetFit,testing[,-1])
```

3. Run an ensemble model for regression using the **caretEnsemble** or **mlr** package and compare the results with the elastic net and random forest model. Did the test accuracy increase?
**HINT:** You need to install these extra packages and learn how to use them in the context of ensemble models. [Difficulty: **Advanced**] 

**solution:**
```{r}
library(caretEnsemble)

ensemblList <- caretList(Age ~., data = trainingData,
                        trControl = ctrlNet,
                        methodList = c("glm", "lm")
                        )

modelCor(resamples(ensemblList))
```

```{r}
ensembleModel <- caretEnsemble(ensemblList, 
                              metric = "accuracy",
                              summaryFunction = twoClassSummary,
                              classProbs = TRUE
                              )
summary(ensembleModel)
```

